Generalizing the Model View Controller Paradigm

### 2006-08-09 inst/paper/paper2v2.tex
Added new versions of papers and added paper 2 version 2 (paper2vs.tex) that is my current version to give to the committee.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@19090 bc3139a8-67e5-0310-9ffc-ced21a209358

\subsection{Background}\label{Ssec:Backg}

There are many ways to visualize multivariate data where more than
two or three dimensions are needed to view the data.  Several of these ideas
were discussed in previous work \cite{EW05}, so the focus in this paper
is on one specific method of visualizing multivariate data that is
particularly useful for exploratory data analysis; creating linked,
interactive views of linked data sets.

First definitions of linked views, interactive views and linked data
sets are given in the following paragraphs.  Linked views mean that if a
component that represents data (such as a point on a plot or a row in a
spreadsheet) changes its appearance on one view, then the corresponding
component on a second view also changes its appearance.  For components on
different views to be corresponding, the data displayed in those components
must come from the same entity in a model.  As an example, suppose that the
model consists of the gene expression levels from four samples.  If the same
one hundred genes are studied in the four samples, then the model could be
represented as a matrix of four columns by one hundred rows with four columns
for the four samples and one hundred rows for the one hundred genes.  In this
model, the entities are the genes.  If two scatter plots are created as views
of this model, where the first plot displays sample one versus sample two and
the second plot displays sample three versus sample four, then the points on
each plot that referred to the same gene are corresponding components.  Linked
views, which have a long history in statistics, are now considered a standard
feature of interactive data visualization software \cite{GGobi}.  Linked views
have been implemented in several pieces of statistical software, such as
XLisp-Stat \cite{Lisp} and GGobi \cite{GGobiMan}.  Implementing linked views
is based on the model-view-controller (MVC) design \cite{DesignPatterns},
which is a widely used and well understood paradigm.  The MVC design is
discussed in more detail in Section \ref{Ssec:Design}.

@@

\begin{figure}[!h]
  \begin{center}    
    \scalebox{0.55}{\includegraphics{Event.jpg}}
    \caption{ Flow Chart of the Response to an Event. }
    \label{Fig:Event}
  \end{center}
\end{figure}

Linked data sets mean that the data have more than one conceptual grouping,
but there is a relationship between these different conceptual groupings.
Linked data sets are an idea that is familiar to users of relational
databases.  Data in a relational database are typically stored in many
linked tables with rows in different tables linked to each other through a
unique identifier called a key.  In this example, each table is
considered a data set and they are linked through the keys.  Another situation
where linked data sets occur is when there is experimental or study data that
are linked to meta-data.  An example of this situation is given in Section
\ref{Ssec:Limit} where the experimental data are microarray data, which looks
at gene expression levels, and the meta-data are a Gene Ontology graph that
gives the molecular functions of the genes studied in the microarray
experiment.

@@

As a motivating example of linked views based on linked data sets, consider
Figure \ref{Fig:firstMP}.  In this figure, there are two views, one view of
data set 1 and one view of data set 2.  Data set 1 and data set 2 are linked.
When a user initially interacts with the scatter plot view of data set 2, the
response to this interaction is to tell data set 2 to update itself (this is
indicated on the figure as a circled 1).  Then in Step 2, data set 2 tells the
scatter plot view that is should update itself in response to the change in the
data.  Next, in Step 3, because data set 2 has changed and because it is
linked to data set 1, data set 2 sends information to data set 1 about the
change.  In Step 4, the information from data set 2 is converted into
information that data set 1 can understand.  In Step 5, this information is
received by data set 1 and data set 1 updates its data.  Finally, in Step 6,
data set 1 tells its view to update itself.  This example shows the components
and the communication between the components that are necessary to create
linked views of underlying linked data sets.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.55}{\includegraphics{firstMessagePassing.jpg}}
    \caption{ Linked views of linked Data Sets.  When a user interacts with
      the y versus x scatter plot, the first step, indicated by a circled one,
      is to tell data set 2 to update.  Then all views of data set 2 must be
      updated.  Next any data that are linked to data set 2 must be notified
      that data set 2 changed.  This information from data set 2 must be
      altered so that it can be understood by linked data sets, which is shown
      in the fourth step.  Finally, data set 1 is notified that it should
      update in response to the change in data set 2 and then all views of
      data set 1 are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

In Section \ref{Ssec:Design}, the design of linked, interactive views based on
the MVC paradigm is discussed.  However, the limitation with this design is
that it expects one data set, which limits the type of data we can visualize.
There are many situations where multiple data sets occur, particularly any
time there are experimental or study data linked to meta-data.  Being able to
visualize data from multiple data sets is helpful to understand the
relationships between variables.  Thus, after discussing the limitations of
the MVC design in Section \ref{Ssec:Limit}, an extension of the MVC design is
discussed in Section \ref{Sec:Extend}.  This extension supports the creation of
linked \Robject{MVC} objects.  The design of individual \Robject{MVC} objects
is discussed in 
Section \ref{Sec:OneMVC} while the implementation of linked \Robject{MVC}
objects is discussed in Section \ref{Sec:MultMVC}.  The goals of the software
discussed in this paper are to create linked views of linked data sets, to
create interactive views where the response to an event can be changed, and to
create an extensible design that allows future users to make additions.

\subsection{Design and Implementation of Linked, Interactive
  Views}\label{Ssec:Design}

Focusing on linked, interactive views as a powerful visualization tool, the
next step is to determine how to design a software package that can implement
this method.  A popular design for creating linked views of data is the
MVC paradigm.  This design consists of three types of objects: the controller,
which defines what actions occur in response to user input; the view, which
consists of displays of the data; and the model, which manages the data.  This
paradigm is powerful because it decouples the views from the model.  Thus,
only one copy of the data needs to be stored and then all views refer to this
data set.  With all views based on the same data, any changes to the data can
be propagated to all views, which creates linked views.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{ExMVC.jpg}}
    \caption{ A Model with its Views. }
    \label{Fig:ExMVC}
  \end{center}
\end{figure}

Notice that although the components (model, view, and controller) in the MVC
design are independent of one other, there needs to be communication
between them so that information can be shared.  Examples of this
communication between components include the model notifying views that the
data have changed, a view asking the controller how it should respond to user
interaction, and a view telling the model that the data should change in
response to user interaction with the view.  This message passing between
components in the MVC paradigm is discussed in detail in Section
\ref{Ssec:OneMess}.

In \cite{EW05}, we discuss the creation of R software to fulfill three goals:
to create views that are linked, to create views that are interactive, and to
create a design that is extensible.  To satisfy the overall goals of the
design, two R packages, called \Rpackage{MVCClass} and \Rpackage{iSPlot},
were created.  The \Rpackage{MVCClass} package implements the class and
generic function definitions needed for the MVC design and the
\Rpackage{iSPlot} package creates the front end to let the user create linked,
interactive views of two-dimensional data, such as data frames and matrices,
through a GUI or through command line functions.  The reason for
separating the software into two packages was that it allowed the class
definitions in the \Rpackage{MVCClass} package to be used by any other R
package that wanted to create linked views.

\subsection{Limitations of the MVC Design}\label{Ssec:Limit}

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{newDatabaseTables.jpg}}
    \caption{ Example of Relational Database Tables. }
    \label{Fig:DBTab}
  \end{center}
\end{figure}

Clearly, this information could be combined into one large table, called
CombinedStateCity, as shown in Figure \ref{Fig:OneDBTab}.  The
CombinedStateCity table has the following columns: State, Life
Expectancy, Income, City, Precipitation, and Temperature.  Now if a state's
life expectancy changes, all instances of this state need to change
in the new large table.  For example, if the state had the information for
three cities recorded, then the state has three rows in the
CombinedStateCity table and each row needs to change the value for the
state's life expectancy.  If instead only one record was kept per state, as
shown in Figure \ref{Fig:DBTab} above, then this process is less error
prone and requires less memory to store the data.  Another consideration in
combining data sets is that it may give the appearance of a relationship
between variables that does not exist.  For example, it may not make sense to
compare the Income data stored in the State table to the Precipitation data
stored in the City table.  Combining the data tables assumes that there is a
relationship between the variables in all of the tables.  This example shows
that there are advantages to storing the data in separate structures.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{newOneDatabaseTable.jpg}}
    \caption{ Example of a Combined Table. }
    \label{Fig:OneDBTab}
  \end{center}
\end{figure}

In a relational database all of the data are stored in tables, but consider the
situation where there are multiple data sets that are of different data
structures.  For example, suppose that you have experimental microarray data,
which look at gene expression levels.  This experimental data can be
stored as a two dimensional data structure, such as a matrix, where rows are
genes and columns are samples.  Linked to this experimental data are meta-data
that give information about the genes studied in the microarray experiment.
For example, this meta-data may be terms from Gene Ontology, which form a
directed acyclic graph, describing the molecular functions carried out by
these genes.  These two data sets (the experimental data and the meta-data)
are linked through the genes.  Each gene can have more than one molecular
function and each molecular function can have more than one gene annotated at
that function so there is a many-to-many relationship between the genes and
the molecular functions.  Though these two data sets are linked, the
microarray data and the Gene Ontology data are different data structures.

Now for the original MVC design, where one data set is expected, it is not
clear how to combine these two data sets into one data set.  The meta-data
graph could be stored as a matrix, but a graph is a data structure that can be
well represented as an object with slots for nodes and edges, especially if
the graph is sparse.  As with relational database tables, combining the data
sets complicates data storage, alteration, and retrieval.  For example, by
combining the data sets a structure may be forced on the data that may not fit
well, especially if the original data sets are instances of different classes.
In contrast, keeping the data sets stored separately in their original
structures simplifies the processes of storing, updating, and accessing the
data.

\section{Extending the MVC Design}\label{Sec:Extend}

As discussed in Section \ref{Sec:Intro}, the MVC design typically uses a single
conceptual unit, such as a data frame or a matrix, for the model.  In Section
\ref{Ssec:Limit} we provided several examples where it is advantageous to link
multiple data sets.  These examples included linked tables in relational
databases and experimental data linked to meta-data.  Being able to visualize
variables simultaneously from different data sets that are linked is useful for
exploring relationships between these variables.  We next describe how the MVC
design can be revised and extended to accommodate multiple data sets.

\subsection{One Large Data Set}\label{Ssec:OneDS}

One possibility for dealing with multiple data sets within the MVC design is to
combine the multiple data sets into one large data set.  With this new
combined data, the original MVC design can be used to create linked views.

Unfortunately, as discussed in Section \ref{Ssec:Limit}, combining
multiple data sets into one large data set can be problematic.  Even if the
multiple data sets have the same data structure, combining the data sets often
results in redundant data and makes updating and representing the
data more complicated.  In the situation where the data sets do not have the
same structure, it may be necessary to define a new class to appropriately
represent the combined data.  Also, combining the data into one data structure
may give the appearance that some variables have a relationship that does not
exist.  Just because two data sets are linked does not necessarily mean that
there is a meaningful relationship between their data.  Thus, another way
of linking views based on multiple data sets needs to be designed.

\subsection{Multiple MVC Objects}\label{Ssec:MMVC}

A more straightforward approach may be to link \Robject{MVCs}.  Each
\Robject{MVC} is a coherent object and the interactions between \Robject{MVCs}
are encapsulated in a set of messages.  We use one \Robject{MVC} object to
handle each data set.  Thus, within a \Robject{MVC} object, one can create
linked views of that particular data set, using the standard MVC design.  But
between \Robject{MVC} objects we design a second messaging system so that a
change in one data set is sent to the other linked data sets.

@@

\begin{table}[h]
  \begin{center}
    \begin{tabular}{ | r | r | }
      \hline
      State & City \\ \hline
      Pennsylvania & Philadelphia \\ \hline
      Pennsylvania & Pittsburgh \\ \hline
      California & Los Angeles \\ \hline
      California & San Francisco \\ \hline
      California & Sacramento \\ \hline
      Minnesota & Duluth \\ \hline
      Minnesota & Minneapolis St. Paul \\ \hline
    \end{tabular}
    \caption{State and City Values.}\label{Tab:CityState}
  \end{center}
\end{table}

\normalsize
As an example, suppose that one has two data sets: one with information at the
state level and one with information at the city level.  These tables are
shown in Table \ref{Tab:CityState}.  The State data include Life Expectancy
(in years from 1970) and Income (per capita from 1974) data
while the City data include Precipitation (in inches from 1975) and Average
High Temperature in January (in Fahrenheit from 2005)
data.  These two data sets are linked by the StateCity table that links cities
with the states they reside in.  Now for these two data sets there are
two \Robject{MVC} objects.  One \Robject{MVC} object for the State data and
its views and one \Robject{MVC} object for the City data and its views.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{MultipleMVC.jpg}}
    \caption{ Example of Two \Robject{MVC} Objects.  These two \Robject{MVC}
      objects are linked by the StateCity table and the color of the points
    indicates the linking between the data.  The points referring to the state
    of Pennsylvania are red, the points referring to the state of California
    are blue, and the points referring to Minnesota are green. }
    \label{Fig:MultMVC}
  \end{center}
\end{figure}

Figure \ref{Fig:MultMVC} shows one view for the State data and one view
for the City data.  You can see how the State and City data are linked based
on the coloring of the points in the plots.  The points referring to the state
of Pennsylvania are red, the points referring to California are blue, and the
points referring to Minnesota are green.  Even though there are two
\Robject{MVC} objects here, with one \Robject{MVC} object for each data set,
it is still clear that the views are linked through the data sets and that is
shown in this example through the coloring.

@@

\section{Design Details of One MVC Object}\label{Sec:OneMVC}

\subsection{Implementation Details}\label{SSec:OneOver}

>>>>

%\subsection{Background}\label{Ssec:Backg}

%There are many ways to visualize multivariate data where more than
%two or three dimensions are needed to view the data.  Several of these ideas
%were discussed in previous work \cite{EW05}, so the focus in this paper
%is on one specific method of visualizing multivariate data that is
%particularly useful for exploratory data analysis; creating linked,
%interactive views of linked data sets.

There are many ways to visualize multivariate data when more than
two or three dimensions are needed to view the data.  Many authors
\citep{intGrUnwin, DynGraphics, IEEEVisual, GGobi, DataDesk}
have suggested that linked, interactive views of the data
provide a valuable and important set of tools for comprehending such
data.  The model view controller (MVC) paradigm is now a standard
mechanism for implementing such systems.  The paradigm has proven very
effective and allows for clear separation of different components.
In this paper we consider an extension of the MVC paradigm to allow
for, what we believe to be, an easy expansion to linked data sets.

The sorts of problems we are interested in, are those where a user
might begin with a large data set, identify some interesting subset
and want to create interactive views of that subset.  In some cases the
natural data representation of the subset is in a form other than that
of the initial data set.  This leads to complications in the usual MVC
paradigm, where the data (i.e. the model) are often presumed to be in
a single, standard format.  We found that adding new views under such
conditions could be quite problematic and often led to very complex
code.  As we shall demonstrate, there are also issues that arise due to
mappings that are not one-to-one between elements on different views.
Hence we considered an approach that created new instances of
the \Robject{MVC} that were linked to the parent that spawned them.

In particular, our implementation consists of a hierarchy of \Robject{MVCs},
stored as a tree.  Each parent can have multiple children, but the
children have a single parent.  Communication between parent and
child is bidirectional, and a variety of different filters can be used
to control flow.  Thus, we can devise a number of purpose built \Rclass{MVC}
classes, each with a model appropriate to the specific data of
interest, that can be linked in a natural way.  Interacting with the view
in a child \Robject{MVC}, first sends a message to the child \Robject{MVC},
which will update all of its views, but the message will also propagate to the
parent \Robject{MVC} (and further up to the root).  The parent \Robject{MVC}
can choose whether to react to messages from its children -- all control being
in the hands of the users, should they want it.

\subsection*{A motivating example}
\label{sec:micro}

Our development of this model was driven primarily by an interest in
analyzing high throughput biological data.  Such data tend to be large
and very complex, and hence ideally suited to interactive
visualization.  In order to provide the reader with some idea of the
sorts of interactions we would like to support, we provide here a short
description of a type of experiment and some of the likely
interactions that would be of interest to a data analyst.

A gene expression experiment provides data on the expression of mRNA
abundance on thousands to tens of thousands of genes for some number
of samples.  Usually there are many fewer samples than genes.  There is
much that is known about the genes, such as what chromosome they are
located on and what sorts of activities they are engaged in, and in many
cases we want to restrict attention to a specific subset of genes, or
to alter the view so that rather than being gene centric it becomes
centered on some biological aspect, such as chromosome location.

One particularly rich data source is the Gene Ontology (GO), \cite{GO}
which provides three controlled vocabularies that are represented as
directed acyclic graphs, with a root node.  Terms typically have a
child-parent relationship, with the child terms being more specific
and the parent terms more general.  The root node represents the most
general term.  Genes are then annotated at specific nodes in these
graphs where the node is determined, by a domain expert, to describe a
property of the gene.  A gene is then also associated with all parents,
since if the specific term applies so must all more general ones.

There is often substantial interest in relating gene expression data
to such a graph based model, or to chromosomal locations, or to
activity in known biological pathways. Data mappings for this latter
aspect can be obtained from different sources such as \cite{KEGG}.
Further, users typically want to begin with a heatmap view. Then from
that derive a GO graph view, and additionally a chromosomal location
view. In order to interact with the GO graph, the primary unit is the
node in that graph (and sometimes one expects with the edges), while
for the heatmap view the basic element is generally the gene.  In Figures
\ref{Fig:heatmap} and \ref{Fig:GOgraph}, a heatmap view of gene expression
from Chronic Lymphocytic Leukemia (CLL) patients and the corresponding GO
graph are shown, respectively.  By interacting with the heatmap to color one
gene blue (shown as gene PDE8A on the heatmap), which causes the corresponding
GO nodes to be colored blue, one can easily see how this gene relates to the
GO terms.  Also by interacting with the node with an arrow next to it in
Figure \ref{Fig:GOgraph}, which represents the signal transducer activity GO
term, one can create a heatmap view that shows only the genes annotated at
that node, as shown in Figure \ref{Fig:heatmapsub}.  Notice how all genes
annotated at the signal transducer activity GO term have a blue label in
Figure \ref{Fig:heatmapsub}, which shows that this view was derived from a
blue colored node in Figure \ref{Fig:GOgraph}.  Compare the heatmaps in
Figures \ref{Fig:heatmapsub} and \ref{Fig:heatmap}.  The heatmap in Figure
\ref{Fig:heatmapsub} is just a subset of the heatmap in Figure
\ref{Fig:heatmap}, however in Figure \ref{Fig:heatmapsub} eleven genes have a
blue label and in Figure \ref{Fig:heatmap} only one gene has a blue label.
This difference in these heatmaps indicates that views are synchronized based
on the parent-child relationships.  Figure \ref{Fig:heatmap} is the parent of
Figure \ref{Fig:GOgraph}, which is the parent of Figure \ref{Fig:heatmapsub}.
Figures \ref{Fig:heatmap} and \ref{Fig:GOgraph} are synchronized, and Figures
\ref{Fig:GOgraph} and \ref{Fig:heatmapsub} are synchronized, but Figures
\ref{Fig:heatmap} and \ref{Fig:heatmapsub} are not synchronized by color
because they do not have a parent-child relationship: Figure \ref{Fig:heatmap}
is the grandparent of Figure \ref{Fig:heatmapsub}.

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=5in, width=5in]{heatmap.jpg}
    \caption{ Heatmap view of gene expression from Chronic Lymphocytic
      Leukemia (CLL) patients. }
    \label{Fig:heatmap}
  \end{center}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=4.3in, width=4.3in]{GOgraph2.jpg}
    \caption{ Gene Ontology (GO) graph that is associated with the genes
      displayed in the heatmap in Figure \ref{Fig:heatmap}.  Notice that one
      gene (PDE8A) is colored blue in the heatmap and all of the corresponding
      GO terms are colored blue in the graph.  The arrow next to one of the
      nodes indicates that this node was used to create the view shown in
      Figure \ref{Fig:heatmapsub}.  The node with the arrow next to it
      represents the signal transducer activity GO term. }
    \label{Fig:GOgraph}
  \end{center}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=5in, width=5in]{heatmapsub.jpg}
    \caption{ A heatmap view that shows only the genes that are annotated at
      the node in Figure \ref{Fig:GOgraph} that has an arrow next to it.  This
      node represents the GO term, signal transducer activity, and as shown in
      this heatmap, eleven genes are annotated at this node.  Notice that all
      the labels in this heatmap are colored blue, indicating that the parent
      node that was used to create this view was also colored blue.  
     }
    \label{Fig:heatmapsub}
  \end{center}
\end{figure}

\subsection*{Overview}
\label{Ssec:Over}

We found that constructing a single MVC that would encompass all
anticipated views was very cumbersome. It would also be inefficient
since a MVC model for the GO graph, or for a chromosome view would
also be of substantial interest, regardless of whether there was an
initial heatmap view or not. Thus, we looked for a solution that was
based more on component software constructed within an object oriented
programming environment.
%%FIXME: cite the JSS submission
We had several objectives, or design criteria. They were that the
software should be extensible in a high level language. We chose R as
the language for our implementation and wanted to allow users that
knew no other languages to be able to extend and enhance the
system.  Therefore, we built our software on top the \Rpackage{RGtk} package of
D. Temple Lang, since it provides such an interface. We also vastly
preferred Gtk+ over Tcl/Tk since its widgets are more sophisticated,
which we found results in a much shorter development time.

We wanted to provide a system that was usable and extensible by
relatively naive end users and hence have provided a graphical user
interface (GUI) which they can use to both carry out an analysis, but
also supports a variety of extensions, such as changing the actions
performed in response to events. We also wanted to support a more
sophisticated developer who might want to take our basic infrastructure
and create an interactive system, say for the interactive analysis of
time series data, using our design. Hence, all commands are
available both through the GUI and through command line functions.

\section{Methods}
\label{Sec:Methods}

First definitions of linked views, interactive views and linked data
sets are given in the following paragraphs.  Linked views mean that if a
component that represents data (such as a point on a plot or a row in a
spreadsheet) changes its appearance on one view, then the corresponding
component on a second view also changes its appearance.  For components on
different views to be corresponding, the data displayed in those components
must come from the same entity in a model.  As an example, suppose that the
model consists of the gene expression levels from four samples.  If the same
one hundred genes are studied in the four samples, then the model could be
represented as a matrix of four columns by one hundred rows with four columns
for the four samples and one hundred rows for the one hundred genes.  In this
model, the entities are the genes.  If two scatter plots are created as views
of this model, where the first plot displays sample one versus sample two and
the second plot displays sample three versus sample four, then the points on
each plot that referred to the same gene are corresponding components.  Linked
views, which have a long history in statistics, are now considered a standard
feature of interactive data visualization software \cite{GGobi}.  Linked views
have been implemented in several pieces of statistical software, such as
XLisp-Stat \cite{Lisp}, iPlots
\cite{iPlots}, and GGobi \cite{GGobiMan}.  Implementing linked views
is based on the model-view-controller (MVC) design \cite{DesignPatterns},
which is a widely used and well understood paradigm.  The MVC design is
discussed in more detail in Section \ref{Ssec:Design}.

@@

\begin{figure}[!h]
  \begin{center}    
    \scalebox{0.55}{\includegraphics{Event.jpg}}
    \caption{ Flow chart of the response to an event. }
    \label{Fig:Event}
  \end{center}
\end{figure}

Linked data sets mean that the data have more than one conceptual grouping,
but there is a relationship between these different conceptual groupings.
Linked data sets are an idea that is familiar to users of relational
databases.  Data in a relational database are typically stored in many
linked tables with rows in different tables linked to each other through a
unique identifier called a key.  In this example, each table is
considered a data set and they are linked through the keys.  Another situation
where linked data occur is when there is experimental or study data that
are linked to meta-data.  An example of this situation is given in Section
\ref{Ssec:Limit} where the experimental data are microarray data, which looks
at gene expression levels, and the meta-data are a Gene Ontology graph that
gives the molecular functions of the genes studied in the microarray
experiment.

@@

Consider Figure \ref{Fig:firstMP}, which provides a cartoon
description of the steps that occur when a user interacts with a
view.  In the cartoon there are only two views, one of model A (note 
that once a data set is loaded in a MVC, we call it a model) and the
other of the linked model B.  When a user interacts with the scatter plot view
of model B, the response to this interaction is for model B to update itself
(this is indicated on the figure as a circled 1).  In Step 2, the scatter plot
view is told to update itself.  Next, in Step 3, because model B is linked
to model A, model B sends a message to model A.  In Step 4,
the information from model B is converted into information that
model A understands.  In Step 5, this information is received by
model A and model A updates itself.  Finally, in Step 6, model
A tells its view to update itself.  This example shows the components
and the communication between the components that are necessary to create
linked views of underlying linked data.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.55}{\includegraphics{firstMessagePassing.jpg}}
    \caption{ Linked views of linked models.  When a user interacts with
      the scatter plot, the first step, indicated by a circled 1,
      is to tell model B to update.  Then all views of model B must be
      updated.  Next any data that are linked to model B must be notified
      that model B changed.  This information from model B must be
      altered so that it can be understood by linked models, which is shown
      in the fourth step.  Finally, model A is notified that it should
      update in response to the change in model B and then all views of
      model A are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

In Section \ref{Ssec:Design}, the design of linked, interactive views based on
the MVC paradigm is discussed.  However, the limitation with this design is
that it expects one data set, which limits the type of data we can visualize.
Thus, after discussing the limitations of
the MVC design in Section \ref{Ssec:Limit}, an extension of the MVC design is
discussed in Section \ref{Sec:Extend}.  This extension supports the creation of
linked \Robject{MVC} objects.  The design of individual \Robject{MVC} objects
is discussed in Section \ref{Sec:OneMVC} while the implementation of linked
\Robject{MVC} objects is discussed in Section \ref{Sec:MultMVC}.  The goals of
the software discussed in this paper are to create linked views of linked data
sets, to create interactive views where the response to an event can be
changed, and to create an extensible design that allows future users to make
additions.

\subsection{Design and Implementation of Linked, Interactive Views}
\label{Ssec:Design}

Focusing on linked, interactive views as a powerful visualization tool, the
next step is to determine how to design a software package that can implement
this method.  A popular design for creating linked views of data is the
MVC paradigm.  This design consists of three types of objects: the controller,
which defines what actions occur in response to user input; the view, which
consists of displays of the data; and the model, which manages the data.  This
paradigm is powerful because it decouples the views from the model.  Thus,
only one copy of the data need to be stored and then all views refer to this
data set.  With all views based on the same data, any changes to the data can
be propagated to all views, which creates linked views.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{ExMVC.jpg}}
    \caption{ A model with its views. }
    \label{Fig:ExMVC}
  \end{center}
\end{figure}

Notice that although the components (model, view, and controller) in the MVC
design are independent of one other, there needs to be communication
between them so that information can be shared.  Examples of this
communication between components include the model notifying views that the
data have changed, the controller asking the view which object on the plot the
user just interacted with, and the controller telling the model that the data
should change in response to user interaction with the view.  This message
passing between components in the MVC paradigm is discussed in detail in
Section \ref{Ssec:OneMess}.

In \cite{EW05}, we discuss the creation of \textsf{R} software to fulfill
three goals: to create views that are linked, to create views that are
interactive, and to create a design that is extensible.  To implement the
overall goals of the design, two \textsf{R} packages, called
\Rpackage{MVCClass} and \Rpackage{iSPlot}, were created.  The
\Rpackage{MVCClass} package implements the class and generic function
definitions needed for the MVC design and the \Rpackage{iSPlot} package
creates the front end to let the user create linked, interactive views of
two-dimensional data, such as data frames and matrices, through a GUI or
through command line functions.  The reason for separating the software into
two packages was that it allowed the class definitions in the
\Rpackage{MVCClass} package to be used by any other \textsf{R} package that
wanted to create linked views.

\subsection{Limitations of the MVC Design}
\label{Ssec:Limit}

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{newDatabaseTables.jpg}}
    \caption{ Example of relational database tables. }
    \label{Fig:DBTab}
  \end{center}
\end{figure}

Clearly, this information could be combined into one large table, called
CombinedStateCity, as shown in Figure \ref{Fig:OneDBTab}.  The
CombinedStateCity table has the following columns: State, Life
Expectancy, Income, City, Precipitation, and Temperature.  Now if a state's
life expectancy changes, all instances of this state need to change
in the new large table.  For example, if the state had the information for
three cities recorded, then the state has three rows in the
CombinedStateCity table and each row needs to change the value for the
state's life expectancy.  If instead only one record was kept per state, as
shown in Figure \ref{Fig:DBTab} above, then this process is less error
prone and requires less memory to store the data.  Another consideration in
combining data is that it may give the appearance of a relationship
between variables that does not exist.  For example, it may not make sense to
compare the Income data stored in the State table to the Precipitation data
stored in the City table.  Combining the data tables assumes that there is a
relationship between the variables in all of the tables.  This example shows
that there are advantages to storing the data in separate structures.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{newOneDatabaseTable.jpg}}
    \caption{ Example of a combined table. }
    \label{Fig:OneDBTab}
  \end{center}
\end{figure}

In a relational database all of the data are stored in tables, but consider the
situation where there are linked data that are of different data
structures.  For example, suppose that you have experimental microarray data,
which look at gene expression levels.  This experimental data can be
stored as a two dimensional data structure, such as a matrix, where rows are
genes and columns are samples.  Linked to this experimental data are meta-data
that give information about the genes studied in the microarray experiment.
For example, this meta-data may be terms from Gene Ontology, which form a
directed acyclic graph, describing the molecular functions carried out by
these genes.  The experimental data and the meta-data
are linked through the genes.  Each gene can have more than one molecular
function and each molecular function can have more than one gene annotated at
that function so there is a many-to-many relationship between the genes and
the molecular functions.  Though the data are linked, the
microarray data and the Gene Ontology data are different data structures.

Now for the original MVC design, where one grouping of the data is expected,
it is not clear how to combine the experimental data and meta-data.  The
meta-data graph could be stored as a matrix, but a graph is a data structure
that can be well represented as an object with slots for nodes and edges,
especially if the graph is sparse.  As with relational database tables,
combining the data complicates data storage, alteration, and retrieval.  For
example, by combining the data, a structure may be forced on the data that may
not fit well, especially if the original data are instances of different
classes.  In contrast, keeping the data stored separately in their original
structures simplifies the processes of storing, updating, and accessing the
data.

\section{Extending the MVC Design}
\label{Sec:Extend}

As discussed in Section \ref{Sec:Intro}, the MVC design typically uses a single
conceptual unit, such as a data frame or a matrix, for the model.  And
we believe that such an approach is a good one.  MVCs should be
specialized to certain types of data, as this allows them to provide
better and more natural interactions.  However, we additionally argue
that there is great benefit to considering linking these different
MVCs when the underlying data themselves are derived from the same source.
We believe that there will often be a hierarchical relationship
between the models and the MVCs that contain them.  We have found
that allowing the user to create new MVCs based on a current MVC is an
important and useful paradigm.

Each \Robject{MVC} is a coherent object and the interactions between
\Robject{MVCs} are encapsulated in a set of messages.  We use one
\Robject{MVC} object to handle each model.  Thus, within a
\Robject{MVC} object, one can create linked views of that particular
model, using the standard MVC design.  Between \Robject{MVC}
objects we propose a second messaging system that allows messages to
be sent between linked models.

%In Section
%\ref{Ssec:Limit} we provided several examples where it is advantageous to link
%multiple data sets.  These examples included linked tables in relational
%databases and experimental data linked to meta-data.  Being able to visualize
%variables simultaneously from different data sets that are linked is useful
%for exploring relationships between these variables.  We next describe how
%the MVC design can be revised and extended to accommodate multiple data sets.

%\subsection{One Large Data Set}\label{Ssec:OneDS}
%
%One possibility for dealing with multiple data sets within the MVC design is
%to combine the multiple data sets into one large data set.  With this new
%combined data, the original MVC design can be used to create linked views.
%
%Unfortunately, as discussed in Section \ref{Ssec:Limit}, combining
%multiple data sets into one large data set can be problematic.  Even if the
%multiple data sets have the same data structure, combining the data sets often
%results in redundant data and makes updating and representing the
%data more complicated.  In the situation where the data sets do not have the
%same structure, it may be necessary to define a new class to appropriately
%represent the combined data.  Also, combining the data into one data structure
%may give the appearance that some variables have a relationship that does not
%exist.  Just because two data sets are linked does not necessarily mean that
%there is a meaningful relationship between their data.  Thus, another way
%of linking views based on multiple data sets needs to be designed.

%\subsection{Multiple MVC Objects}\label{Ssec:MMVC}

@@

\begin{table}[h]
  \begin{center}
    \begin{tabular}{ | r | r | }
      \hline
      State & City \\ \hline
      Pennsylvania & Philadelphia \\ \hline
      Pennsylvania & Pittsburgh \\ \hline
      California & Los Angeles \\ \hline
      California & San Francisco \\ \hline
      California & Sacramento \\ \hline
      Minnesota & Duluth \\ \hline
      Minnesota & Minneapolis St. Paul \\ \hline
    \end{tabular}
    \caption{State and city values.}\label{Tab:CityState}
  \end{center}
\end{table}

\normalsize
As an example, suppose that one has two models: one with information at the
state level and one with information at the city level.  These tables are
shown in Table \ref{Tab:CityState}.  The State data include Life Expectancy
(in years from 1970) and Income (per capita from 1974) data
while the City data include Precipitation (in inches from 1975) and Average
High Temperature in January (in Fahrenheit from 2005)
data.  These two models are linked by the StateCity table that links cities
with the states they reside in.  Now for these two models there are
two \Robject{MVC} objects.  One \Robject{MVC} object for the State data and
its views and one \Robject{MVC} object for the City data and its views.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{MultipleMVC.jpg}}
    \caption{ Example of two \Robject{MVC} objects.  These two \Robject{MVC}
      objects are linked by the StateCity table and the color of the points
    indicates the linking between the data.  The points referring to the state
    of Pennsylvania are red, the points referring to the state of California
    are blue, and the points referring to Minnesota are green. }
    \label{Fig:MultMVC}
  \end{center}
\end{figure}

Figure \ref{Fig:MultMVC} shows one view for the State data and one view
for the City data.  You can see how the State and City data are linked based
on the coloring of the points in the plots.  The points referring to the state
of Pennsylvania are red, the points referring to California are blue, and the
points referring to Minnesota are green.  Even though there are two
\Robject{MVC} objects here, with one \Robject{MVC} object for each model,
it is still clear that the views are linked through the models and that is
shown in this example through the coloring.

@@

\section{Design Details of One MVC Object}
\label{Sec:OneMVC}

\subsection{Implementation Details}
\label{SSec:OneOver}
